\chapter{Obtaining the Project}

A short demo of the project can be found on
\url{http://www.youtube.com/watch?v=h9gecC5woxs} or the root of the accompanying
CD.

The source code for this master's thesis is also located on the accompanying CD
in the folder RayTracingDynamicScenes. The application developed for this thesis
has been produced in the open source 3D engine framework OpenEngine and I have
therefore not implemented everything found in RayTracingDynamicScenes. The
implementation of my ray tracer can be found under the project/RayTracing
folder, which contains the scene setup, and under extensions/PhotonMapping,
where the code for construction kd-trees and ray tracing them is located. The
project has been compiled and tested on Max OS X and Ubuntu.

If the CD is not present, then the project can be obtained by installing
OpenEngine and then installing the RayTracing project. OpenEngine requires
Python, CMake, Boost and the version control system Darcs to be installed on the
system. My project additionally requires that SDL, CUDA 3.1 and FreeImage be
installed on the system.

To install OpenEngine simply run the following command in a terminal

\terminalCommand{darcs get http://openengine.dk/code/openengine}

This will place the OpenEngine source in a folder named \textit{openengine} in
the current working directory. cd into that directory and execute the commands

\terminalCommand{chmod +x dist.py}

\terminalCommand{chmod +x make.py}

The project and its dependencies can then be installed with the command

\terminalCommand{./dist.py install proj:RayTracing}

The project can be compiled by running

\terminalCommand{./make.py}

and executed with the command

\terminalCommand{./build/RayTracing/RayTracing}
